---
title: "Fundamentals of R"
subtitle: "Linear Regression"
author: "Prepared By Clinton Moshe Mwachia"
description: "Linear Modelling in R"
date: "`r Sys.Date()`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(tidyverse)
library(gt)
library(mice)
library(caret)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Machine learning is a powerful approach to extract insights and make predictions from data. In this lesson, we will explore how to implement `machine learning` using `linear models` in R. 

We will cover the entire process, from loading and cleaning the data to model training, evaluation, and making predictions.

The lesson aims to predict the sale price of a house using other features provided.

*NOTE: THE STEPS PROVIDED HERE ARE NOT THE ONLY WAY OF DOING ML, YOU SHOULD TRY DIFFERENT APPROACHES AND THEN CHOOSE THE BEST FOR YOUR DATA.*

## Data Loading

The lesson uses the `House Prices - Advanced Regression Techniques` data obtained from **Kaggle** website. The data is split into train set, used for training the model and test set, used for testing the model.

```{r data-loading}
#setwd("F:/R/RMastery/inst/tutorials/linear_regression")
# loading the training data
train = read.csv("data/train.csv")

# see all columns of the data
summary(train)
```

The data has 81 variables and 1460 observations, we will use the variables: *SalePrice*, *YrSold*, *GarageArea*, *GarageYrBlt*, *YearBuilt*, *LotArea*, and *LotFrontage* for our analysis and modelling.

```{r select-cols}
# the columns
cols = c("SalePrice", "YrSold", "LotFrontage", "LotArea", "YearBuilt",
         "GarageArea", "GarageYrBlt")

# filtering the columns
train_updated = train %>%
  select(all_of(cols))
```

## Data Cleaning

Data cleaning is an important step to ensure that the data is in the right format and free from errors or missing values. Cleaning may involve handling missing data, transforming variables, or removing outliers.

**Missing Data**

```{r helper-fn}
# helper function to calculate na percentages of each columns
calc_na = function(data){
  
    na_val = colSums(is.na(data)) / nrow(data)
  
  return(na_val*100)
}
```


```{r missing-data}
# checking missing data
calc_na(train_updated)
```

The variables `Lot Fontage` and `Garage Year Built` contains NA values of 17.7% and 5.5% respectively, the other variables have no missing data.

We will impute the columns with missing data

```{r impute-na}
imp_model <- mice(train_updated, method = "pmm", m = 5)

imputed_data <- complete(imp_model)

# head of the data
head(imputed_data)
```


```{r confirm-na}
# confirming that the imputation worked
calc_na(imputed_data)
```

## Exploratory Data Analysis

To better understand our data, we will perform data visualizations to answer the following questions: -


      1. Are old houses cheaper or more expensive?
      2. Does a large lot area increase the sale price?
      3. Which year has the highest average sale price?
      4. Houses with large garages are more expensive?
      
Let's get started.

**Are old houses cheaper or more expensive?**

To answer this question, we will plot the year build of the house verses the sale price. This will us know the relationship between the year the house was built and the sale price.

```{r saleprice-yearbuilt}
imputed_data %>%
  select(SalePrice, YearBuilt) %>%
  arrange(YearBuilt) %>%
  ggplot(aes(x=YearBuilt, y=SalePrice)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_bw() +
    ggtitle("Sale Price changes by Year of the house")
```

**CONCLUSION:** HOUSES THAT ARE RECENTLY BUILT ARE MORE EXPENSIVE COMPARED TO THOSE THAT ARE BUILD A LONG TIME AGO.

**Does a large lot area increase the sale price?**

```{r saleprice-lotArea}
imputed_data %>%
  select(SalePrice, LotArea) %>%
  ggplot(aes(x=LotArea, y=SalePrice)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_bw() +
    ggtitle("Sale Price changes by house lot Area")
```

**CONCLUSION:** THE SALE PRICE OF A HOUSE INCREASES WITH THE LOT AREA. A HOUSE WITH A LARGE LOT AREA IS EXPENSIVE COMPARED TO A HOUSE WITH A SMALL LOT AREA.

**Which year has the highest average sale price?**

```{r saleprice-yrsold}
imputed_data %>%
  select(SalePrice, YrSold) %>%
  group_by(YrSold) %>%
  summarise(
    avg_salePrice = mean(SalePrice)
  ) %>%
  ggplot(aes(x=YrSold, y=avg_salePrice)) +
    geom_point() +
    geom_line() +
    theme_bw() +
    ylab("Average Sale Price") +
    ylab("Year Sold") +
    ggtitle("House Sale Price average by Year Sold")
```

**CONCLUSION:** HOUSES THAT WERE SOLD ON 2006 AND 2007, HAD THE HIGHEST AVERAGE SALE PRICES AS COMPARED TO THE OTHER YEARS.

**Houses with large garages are more expensive?**

```{r saleprice-GarageA}
imputed_data %>%
  select(SalePrice, GarageArea) %>%
  ggplot(aes(x=GarageArea, y=SalePrice)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_bw() +
    ggtitle("Sale Price changes by Garage Area")
```

**CONCLUSION:** AS THE GARAGE AREA INCREASES, THE SALE PRICE OF THE HOUSE ALSO INCREASES. THIS IMPLIES THAT HOUSES WITH A BIG GARAGE AREA IS MORE EXPENSIVE COMPARED TO A HOUSE WITH A SMALL GARAGE.

## Model 1

**Data Splitting**

```{r data-split}
# split index
# 80% train data, 20% test data
index = createDataPartition(imputed_data$SalePrice, p=0.8,list = FALSE)

# train data
train_data = imputed_data[index,]

# test data
test_data = imputed_data[-index,]
```

**Fitting The Model**

```{r model-1}
# fit the model
lm = lm(SalePrice~., data = train_data)

# model summary
summary(lm)
```

**Model Plots**

```{r model-plots}
par(mfrow=c(2,2))
plot(lm)
```

**Making Predictions**

```{r predictions}
# making the predictions
pred = predict(lm, newdata = test_data)

# Root Mean Squared Error (RMSE) of the MODEL.
rmse <- sqrt(mean((pred - test_data$SalePrice)^2))

rmse
```

The interpretation of what constitutes a "good" value for RMSE depends on the specific context and the scale of the target variable. In general, a lower RMSE indicates better model performance, as it signifies smaller differences between the predicted values and the actual values. However, there is no universally defined threshold for a "good" RMSE value that applies to all situations.

To assess the goodness of an RMSE value, it's helpful to consider the scale and context of the problem you are addressing. You can compare the RMSE to the range of the target variable or the variability in the data. Additionally, it can be beneficial to compare the RMSE of different models or variations of the same model to determine which one performs better.

It's important to note that the notion of a "good" RMSE value is relative and can vary depending on the specific domain, application, and the goals of your analysis. Therefore, it is recommended to interpret the RMSE in conjunction with other evaluation metrics and domain knowledge to make informed decisions about model performance.

## Model 2

**Data Splitting**

```{r data-split-model2}
# scalling the data
scaled_data = imputed_data %>%
  select(-SalePrice) %>%
  scale() %>%
  as.data.frame()

# combining the data
imputed_scaled_data = cbind(scaled_data, SalePrice=imputed_data$SalePrice)

# split index
# 80% train data, 20% test data
index2 = createDataPartition(imputed_scaled_data$SalePrice, p=0.8,list = FALSE)

# train data
train_scaled_data = imputed_scaled_data[index2,]

# test data
test_scaled_data = imputed_scaled_data[-index2,]
```

**Fitting The Model**

```{r model-2}
# fit the model
lm2 = lm(SalePrice~., data = train_scaled_data)

# model summary
summary(lm2)
```

**Model Plots**

```{r model-plots2}
par(mfrow=c(2,2))
plot(lm2)
```

**Predictions**

```{r predictions-model2}
# making the predictions
pred2 = predict(lm2, newdata = test_scaled_data)

# Root Mean Squared Error (RMSE) of the MODEL.
rmse2 <- sqrt(mean((pred2 - test_scaled_data$SalePrice)^2))

rmse2
```

## rmse values

MODEL: `r rmse`

MODEL 2: `r rmse2`

MODEL 2 HAS A MUCH LOWER RMSE COMPARED TO MODEL 1. WE WILL USE MODEL 2 FOR PREDICTIONS

## Final Model Predictions

```{r final-model-predictions}
# loading the test data
test = read.csv("data/test.csv")

# selecting columns of interest
cols2 = c("YrSold", "LotFrontage", "LotArea", "YearBuilt",
         "GarageArea", "GarageYrBlt")

test = test %>%
  select(all_of(cols2))

# SCALING THE VALUES
test_scaled = test %>%
  scale() %>%
  as.data.frame()

# making the predictions
pred = predict(lm2,newdata = test_scaled)
```

CONCLUSION: BASED ON OUR ANALYSIS, MODEL 2 IS THE BEST. THERE ARE MANY WAYS OF IMPROVING THIS MODEL BUT FOR THE SAKE OF THIS LESSON, PLEASE PROCEED TO THE ASSIGNMENT.


## Exercise

### Assignment

FIT TWO OTHER LINEAR MODELS USING DIFFERENT APPROACHES. YOUR TWO MODELS SHOULD BE BETTER THAN THE MODELS WE HAVE ALREADY FITTED. 

1. CREATE A NEW RMD FILE.
2. NAME IT `Linear-Regression-Assignment.Rmd`.
3. COMPLETE THE TASK PROVIDED ABOVE.
4. KNIT TO HTML.

USE APPROPRIATE EVALUATION METHODS.

TIPS:

To evaluate the performance of a linear model and improve its performance, you can consider the following techniques:

1. Residual Analysis:
Residual analysis helps assess how well the linear model fits the data. The residuals are the differences between the actual values and the predicted values. You can plot the residuals against the predicted values to check for any patterns or systematic deviations. A good linear model will have residuals that are randomly scattered around zero.

2. R-squared (Coefficient of Determination):
R-squared measures the proportion of the variance in the dependent variable that can be explained by the independent variables in the model. It ranges from 0 to 1, with higher values indicating a better fit. However, R-squared alone may not provide a complete picture, as it can increase even when irrelevant variables are added to the model. Therefore, it should be interpreted alongside other evaluation metrics.

3. Adjusted R-squared:
Adjusted R-squared adjusts for the number of predictors in the model, penalizing the addition of irrelevant variables. It takes into account both the goodness of fit and the complexity of the model. A higher adjusted R-squared suggests a better trade-off between model complexity and fit.

4. Mean Squared Error (MSE) and Root Mean Squared Error (RMSE):
MSE and RMSE measure the average squared difference between the actual values and the predicted values. These metrics quantify the overall model performance, with lower values indicating better accuracy. They are sensitive to outliers since the differences are squared. Comparing MSE or RMSE values across different models helps identify the model with the smallest errors.

5. Cross-Validation:
Cross-validation is a technique used to estimate the performance of a model on unseen data. It involves splitting the data into multiple subsets, training the model on a portion of the data, and evaluating it on the remaining portion. This process is repeated several times, and the average performance metric is computed. Common cross-validation techniques include k-fold cross-validation and leave-one-out cross-validation.

To improve the performance of a linear model, you can consider the following strategies:

1. Feature Selection:
Identify and select the most relevant features for the model. This can be done through techniques like forward selection, backward elimination, or stepwise regression. Removing irrelevant or redundant features can improve model simplicity and reduce overfitting.

2. Feature Engineering:
Create new features or transform existing ones to capture additional information. This could involve logarithmic or exponential transformations, interaction terms, or polynomial features. Feature engineering can help the model better capture non-linear relationships.

3. Regularization:
Apply regularization techniques such as Ridge regression or Lasso regression. Regularization adds a penalty term to the loss function, discouraging the model from relying too heavily on any single feature. It helps reduce overfitting and improves generalization.

4. Outlier Treatment:
Identify and handle outliers appropriately. Outliers can have a significant impact on the linear model's performance. Depending on the situation, you can consider removing outliers, transforming the data, or using robust regression techniques.

5. Model Comparison and Selection:
Evaluate different linear model variations or alternative models to find the best fit for your data. Consider trying different model assumptions, such as polynomial regression or generalized linear models, to capture complex relationships.

6. Data Quality:
Ensure the data used for modeling is of high quality. Check for missing values, outliers, and inconsistencies in the data. Data cleaning and preprocessing techniques, such as imputation or normalization, can enhance the model's performance.

By employing these evaluation techniques and improvement strategies, you can iteratively refine your linear model to achieve better performance and accuracy. Remember that model evaluation is an ongoing process, and it's essential to assess the model's performance on different datasets and real-world scenarios to ensure its reliability and generalizability.
