---
title: "Fundamentals of R"
subtitle: "Data Structures"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
description: "Data structures in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Data structures define how data is stored and organized in R.

In this lesson, you will learn:

* Vectors
* Lists
* Matrices
* Data Frames
* Bonus: Comments

---

## Variables

Variables in R can be defined using `<-` or `=`.

```{r}
x <- 10
y = 20
```


## VECTORS

A vector stores elements of the same type.

```{r}
vec <- c(1,2,3,4)
vec
```

---

## Exercises

### Exercise 1: Create a Vector

Create a vector called `animals` with 5 animals.

```{r exercise1, exercise=TRUE}
# your code here
```

```{r exercise1-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")
```

```{r exercise1-check}
grade_code()
```

---

### Exercise 2: Vector Length

Find the length of `animals`.

```{r exercise2, exercise=TRUE}
# your code here
```

```{r exercise2-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")

length(animals)
```

```{r exercise2-check}
grade_code()
```

---

### Exercise 3: Indexing

Extract the 1st, 3rd and 5th animals.

```{r exercise3, exercise=TRUE}
# your code here
```

```{r exercise3-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")

animals[c(1,3,5)]
```

```{r exercise3-check}
grade_code()
```

---

## LISTS

Lists can store different data types.

```{r}
my_list <- list(10, TRUE, "R")
my_list
```

---

### Exercise 4: Create a List

Create a list `my_list` using `animals`.

```{r exercise4, exercise=TRUE}
# your code here
```

```{r exercise4-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")

my_list <- list(animals)
```

```{r exercise4-check}
grade_code()
```

---

## MATRICES

Matrices are 2D structures.

```{r}
A <- matrix(c(1,2,3,4,5,6), nrow = 2)
A
```

---

### Exercise 5: Matrix Dimensions

Find dimensions of matrix `A`.

```{r exercise5, exercise=TRUE}
# your code here
```

```{r exercise5-solution}
A <- matrix(c(1,2,3,4,5,6), nrow = 2)

dim(A)
```

```{r exercise5-check}
grade_code()
```

---

### Exercise 6: Matrix Addition

Compute A + A

```{r exercise6, exercise=TRUE}
# your code here
```

```{r exercise6-solution}
A <- matrix(c(1,2,3,4,5,6), nrow = 2)

A + A
```

```{r exercise6-check}
grade_code()
```

---

## DATA FRAMES

Data frames store tabular data.

```{r}
df <- data.frame(one=c(10.5,11,23,4))
df
```

---

### Exercise 7: Summary Stats

Get summary of df

```{r exercise7, exercise=TRUE}
# your code here
```

```{r exercise7-solution}
df <- data.frame(one=c(10.5,11,23,4))

summary(df)
```

```{r exercise7-check}
grade_code()
```

---

### Exercise 8: Add Column

Add a new column called `two`

```{r exercise8, exercise=TRUE}
# your code here
```

```{r exercise8-solution}
df <- data.frame(one=c(10.5,11,23,4))

df$two <- c(1,3,5,7)
```

```{r exercise8-check}
grade_code()
```

---

## BONUS

## Comments

```{r}
# This is a comment
```

---

## FINAL Assignment

### Exercise 9: Full Task

Create:

* animals vector
* marks vector
* list combining both

Then return length of list

```{r exercise9, exercise=TRUE}
# your code here
```

```{r exercise9-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")
marks <- c(70,80,90,85,60)

my_list <- list(animals, marks)

length(my_list)
```

```{r exercise9-check}
grade_code()
```

---

## 🎯 Summary

You learned:

* Vectors
* Lists
* Matrices
* Data Frames
* Basic operations
* Comments
* Variables

---
