---
title: "Fundamentals of R"
subtitle: "R Operations"
author: "Prepared By Clinton Moshe Mwachia"
description: "Perform Various Mathematical Operations in R"
date: "`r Sys.Date()`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Operators are symbols or characters used to perform various mathematical, logical, and relational operations. 

These operators are used to perform calculations, compare values, combine data, and manipulate variables in R. 

Understanding the different types of operators in R is crucial for effective data analysis and programming in R. 

In this context, this note provides an overview of some of the most commonly used operators in R. They include:- 

    1. Assignment operators.
    2. Arithmetic operators.
    3. Comparison operators.
    4. Logical operators.
    5. concatenation operators. 
    
## Arithmetic operators

These operators are used to perform mathematical operations such as addition, subtraction, multiplication, exponentiation, and modulus. These operations are illustrated below.

**Addition**

```{r}
# adding two numbers
2 + 100
```

**Subtraction**

```{r}
# subtracting two numbers
2000 - 100
```

**Multiplication**

```{r}
# multiplying two numbers
44 * 90
```

**Division**

```{r}
# dividing two numbers
1000 / 20
```

**Exponentiation**

```{r}
# exponentiation
2^3
```

**Modulus**

```{r}
# modulus
100 %% 2
# remainder?
100 %% 3
```

## Assignment operators

The assignment operator is used to assign values to variables. You can use either the equal sign or the leftward arrow to assign a value to a variable, both scenarios are illustrated below.

**Using An Equal Sign** 

```{r}
# using an equal sign
x = 10
```

THE VALUE OF X IS `r x`

**Using A leftward arrow** 

```{r}
# using an equal sign
x <- 20
```

THE NEW VALUE OF X IS `r x`

## Comparison operators

These operators are used to compare values. They will always return a Boolean value.

**Equal to**

```{r}
10 == 15
```

**Not Equal to**

```{r}
20 != 15
```

**Less Than or Equal to**

```{r}
2 <= 15
```

**Greater Than or Equal to**

```{r}
20 >= 15
```

**Less Than**

```{r}
100 < 150
```

**Greater Than**

```{r}
100 > 1020
```

## Logical operators

Logical operators are used to perform logical operations.

**Logical AND Operator**

```{r}
TRUE & TRUE

TRUE & FALSE
```

**Logical OR Operator**

```{r}
TRUE | TRUE

TRUE | FALSE
```

**Logical NOT Operator**

```{r}
!TRUE

!(2==5)
```

## concatenation operators

This operator is used to combine variables into a vector.

```{r}
y = c(1,2,3)

y
```

---

### Exercise 1

Create variable *x* with the data *apple, banana, orange* and *y* with the data *rape, pineapple, watermelon*.

Use concatenation operator to combine variable *x* and *y*.


```{r exercise1, exercise=TRUE}
# your code here
```

```{r exercise1-solution}
x <- c("apple", "banana", "orange")

y <- c("grape", "pineapple", "watermelon")

c(x,y)
```

```{r exercise1-check}
grade_code()
```

---

### Exercise 2

Write a function, *calculate_all* that accepts two numbers and performs the following operations:-

    1. Adds the two numbers
    2. Subtracts the second number from the first number
    3. Multiplies the two numbers
    4. Divides the first number by the second number


```{r exercise2, exercise=TRUE}
# your code here
```

```{r exercise2-solution}
calculate_all = function(a, b){
  results <- list(
    Sum = a+b,
    Substraction = a-b,
    Multiplication = a*b,
    Division = ifelse(b != 0, a/b, "Undefined")
  )
  
  return(results)
}

calculate_all(10, 20)
```

```{r exercise2-check}
grade_code()
```

---

### Exercise 3

Using the seq() function, create a vector of integers from 1 to 10. Using arithmetic operators, calculate the sum of the vector.


```{r exercise3, exercise=TRUE}
# your code here
```

```{r exercise3-solution}
nums = seq(1,10,by=1)

sum(nums)
```

```{r exercise3-check}
grade_code()
```

---

### Exercise 4

Create a variable called *age* with a value of 30. Using comparison operators, print "You are an adult" if age is greater than or equal to 18, and "You are not an adult" if age is less than 18.


```{r exercise4, exercise=TRUE}
# your code here
```

```{r exercise4-solution}
age = 30

if(age >= 18){
  cat("You are an adult")
} else {
  cat("You are not an adult")
}
```

```{r exercise4-check}
grade_code()
```

---

### Exercise 5

Using the `rep()` function, create a vector with 10 repetitions of the string "hello". Using the `toUpper()` function, convert the vector to all uppercase letters. Using the concatenation operator, add an exclamation mark to the end of each element in the vector. Print the resulting vector.


```{r exercise5, exercise=TRUE}
# your code here
```

```{r exercise5-solution}
# create the vector
vec = rep("hello", 10)

# convert to upper case
vecUppercase = toupper(vec)

# add exclamation
paste0(vecUppercase,"!")
```

```{r exercise5-check}
grade_code()
```

---