---
title: "Fundamentals of R"
subtitle: "Datasets"
author: "Prepared By Clinton Moshe Mwachia"
description: "Built in and external datasets in R"
date: "`r Sys.Date()`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

R provides a wide range of built-in datasets that are readily available for analysis and exploration.

These datasets cover various domains, including *statistics, economics, biology*, and *social sciences*.

In this lesson, we will explore some commonly used built-in datasets in R.

## In Built Data

R comes with a package called **datasets** that contains numerous built-in datasets.

```         
To access these datasets, we have to load the package. 

To view a list of all the datasets, you can use either 
```

**Method 1**

`data()`

**Method 2**

`library(help = "datasets")`.

After loading the data, you can perform various operations to available datasets.

**View The first 4 rows of `cars` dataset**

```{r}
head(cars, 4)
```

**View The last 4 rows of `cars` dataset**

```{r}
tail(cars, 4)
```

**Get summary statistics of `cars` dataset**

```{r}
summary(cars)
```

## Working Directory

The `working directory` in R refers to the folder or directory where R will look for files and where it will save them by default.

Understanding and managing the working directory is important for efficient file handling and data analysis.

In this section, we will cover the concept of the working directory in R and explore various functions to `set, check, and change` the working directory.

**Get Working Directory**

To check the current working directory in R, you can use `getwd()` function. This function returns the path to the current working directory as a character string.

```{r}
# get current working directory
getwd()
```

**Setting the working Directory**

To set a specific directory as the working directory in R, you can use the `setwd()` function or use the IDE setting. 

This function takes a path to the desired directory as an argument and sets it as the working directory.

```{r}
# set a working directory
#setwd("F:/R/RMastery")
```

**List Files in working directory**

```{r}
list.files()
```

## Writing User Data

In R, users have the flexibility to create and manipulate their own data sets.

This capability allows for the exploration, analysis, and modeling of custom data, tailored to specific needs.

In this lesson, we will cover various techniques for creating our own data in R.

The datasets can be created using the data structures we learned in the `data structures` lesson. They can be created using:

1.  Data Frames.
2.  Vectors.
3.  Matrices.
4.  Lists.
5.  Factors.

```{r}
# create the dataset
data = data.frame(
  name=c("James", "Sydney", "Esther"),
  age = c(23, 34, 67),
  gender = c("M","M","F"))

# to write the data in the current folder, we have to change the working directory
setwd("F:/R/RMastery/inst/tutorials/data")

# create the `data` folder
dir.create("data")

# write csv file
write.csv(data, file = "./data/data.csv")

# write txt file
write.table(data, file = "./data/data.txt")

# write json file
jsonlite::write_json(data, path = "./data/data.json")

# write xlsx file
openxlsx::write.xlsx(data, file = "./data/data.xlsx")

# sqlite
con = RSQLite::dbConnect(drv = RSQLite::SQLite(), dbname = "./data/database.db")

# write data into the db
RSQLite::dbWriteTable(con, "data", data, overwite=TRUE, append=TRUE)
```

## Load External Data

In this section, we will load the datasets we created above.

**Loading CSV DATA**

```{r loading-data-csv}
# loading csv data
data = read.csv("data/data.csv")

# head of the data
head(data)
```

**Loading EXCEL DATA**

```{r loading-data-excel}
# loading csv data
data2 = readxl::read_excel("data/data.xlsx")

# head of the data
head(data2)
```

**Loading TEXT FILE DATA**

```{r loading-data-text}
# loading txt file data
data4 = read.table("data/data.txt", header = TRUE)

# head of the data
head(data4)
```

**Loading JSON DATA**

```{r loading-data-json}
# loading json data
data5 = jsonlite::fromJSON("data/data.json")

# head of the data
head(data5)
```

**Loading SQLITE DATA**

```{r loading-data-sqlite}
# query to read data from the sqlite database
query <- "SELECT * FROM data"

# execute the query
data <- RSQLite::dbGetQuery(con, query)

# head of the data
head(data)

# disconnect the connection
RSQLite::dbDisconnect(con)
```

---

### Exercise 1

Create a vector called `animals` with 5 animals.

```{r exercise1, exercise=TRUE}
# your code here
```

```{r exercise1-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")
```

```{r exercise1-check}
grade_code()
```

---