---
title: "Fundamentals of R"
subtitle: "Probability Distributions"
author: "Prepared By Clinton Moshe Mwachia"
description: "Generate Probability Distributions using R"
date: "`r Sys.Date()`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

Probability distributions are a fundamental concept in statistics and data analysis. In R, there are various built-in functions and packages that allow for the generation and manipulation of probability distributions.

In this lesson, we will explore some commonly used probability distributions in R and demonstrate how to work with them in R.

## Discrete Probability Distributions

**Binomial Distribution**

The binomial distribution models the number of successes in a fixed number of independent Bernoulli trials. In R, the `rbinom()` function generates random numbers from a binomial distribution.

```{r r-binom}
# Generating random numbers from a binomial distribution
n <- 10  # number of trials

p <- 0.5  # probability of success on each trial

binom <- rbinom(n, size = n, prob = p)

binom
```

**Poisson Distribution**

The Poisson distribution models the number of events occurring in a fixed interval of time or space. In R, the `rpois()` function generates random numbers from a Poisson distribution.

```{r r-pois}
# Generating random numbers from a Poisson distribution
lambda <- 2  # average rate of events

poisson <- rpois(n, lambda = lambda)

poisson
```

## Continuous Probability Distributions

**Normal Distribution**

The normal distribution, also known as the Gaussian distribution, is widely used in statistics and data analysis. In R, the `rnorm()` function generates random numbers from a normal distribution.

```{r r-norm}
# Generating random numbers from a normal distribution
mean <- 0  # mean of the distribution

sd <- 1  # standard deviation of the distribution

normal <- rnorm(n, mean = mean, sd = sd)

normal
```

**Uniform Distribution**

The uniform distribution represents a continuous probability distribution where all outcomes are equally likely. In R, the `runif()` function generates random numbers from a uniform distribution.

```{r r-unif}
# Generating random numbers from a uniform distribution
min <- 0  # minimum value

max <- 1  # maximum value

uniform <- runif(n, min = min, max = max)

uniform
```

## Additional Distributions

**Exponential Distribution**

```{r r-exp}
# Generating random numbers from an exponential distribution
rate <- 0.5  # rate parameter

exponential <- rexp(n, rate = rate)

exponential
```

**Gamma Distribution**

```{r r-gamma}
# Generating random numbers from a gamma distribution
rate <- 0.5  # rate parameter

shape <- 0.5  

gamma <- rgamma(n, shape = shape,rate = rate)

gamma
```

**Chi-Square Distribution**

```{r r-chisq}
# Generating random numbers from a chi-square distribution
df <- 5  # degrees of freedom

chi_square <- rchisq(n, df = df)

chi_square
```

---

## Exercises

### Exercise 1

Generate 10 random numbers from a uniform distribution between -1 and 1 in R, and calculate the mean and standard deviation of the generated values.

```{r exercise1, exercise=TRUE}
# your code here
```

```{r exercise1-solution}
# generate the numbers
rnumbers = runif(10, -1, 1)

# mean
mean(rnumbers)

# sd
sd(rnumbers)
```

```{r exercise1-check}
grade_code()
```

---

### Exercise 2

Plot the probability mass function (PMF) of a binomial distribution with n = 50 and p = 0.4 using R's base graphics system.

```{r exercise2, exercise=TRUE}
# your code here
```

```{r exercise2-solution}
# Parameters
n <- 50
p <- 0.4

# Possible values (0 to n)
x <- 0:n

# PMF values
pmf <- dbinom(x, size = n, prob = p)

# Plot using base R
plot(x, pmf, 
     type = "h",              # vertical lines (like a PMF)
     lwd = 2,                 # line width
     col = "blue", 
     xlab = "Number of Successes",
     ylab = "Probability",
     main = "Binomial PMF (n = 50, p = 0.4)")

# Add points for clarity
points(x, pmf, pch = 16, col = "red")
```

```{r exercise2-check}
grade_code()
```

---

### Exercise 3

Calculate the quatiles of a Poisson distribution with λ = 2.5 in R.

```{r exercise3, exercise=TRUE}
# your code here
```

```{r exercise3-solution}
# Parameter
lambda <- 2.5

# Quartiles
quartiles <- qpois(c(0.25, 0.5, 0.75), lambda = lambda)

quartiles
```

```{r exercise3-check}
grade_code()
```

---

### Exercise 4

Calculate the quatiles of a Poisson distribution with λ = 2.5 in R.

```{r exercise4, exercise=TRUE}
# your code here
```

```{r exercise4-solution}
# Parameter
lambda <- 2.5

# Quartiles
quartiles <- qpois(c(0.25, 0.5, 0.75), lambda = lambda)

quartiles
```

```{r exercise4-check}
grade_code()
```

---


### Exercise 5

Generate random numbers from a binomial distribution with parameters n = 30 and p = 0.65 in R and print the numbers.

```{r exercise5, exercise=TRUE}
# your code here
```

```{r exercise5-solution}
# Parameters
n <- 30
p <- 0.65

# Possible values (0 to n)
x <- 0:n

# PMF values
pmf <- dbinom(x, size = n, prob = p)

pmf
```

```{r exercise5-check}
grade_code()
```

---

### Exercise 6

Generate 1000 random numbers from a binomial distribution with parameters n = 30 and p = 0.65 in R and print the numbers. Make the numbers reproducible.

```{r exercise6, exercise=TRUE}
# your code here
```

```{r exercise6-solution}
# Parameters
n <- 30
p <- 0.65

# Generate 1000 random values
set.seed(123)  # for reproducibility
x <- rbinom(1000, size = n, prob = p)

# Print the numbers
x
```

```{r exercise6-check}
grade_code()
```

---

### Exercise 7

Use the values generated from `Exercise 6` to Plot a density plot.

```{r exercise7, exercise=TRUE}
# your code here
```

```{r exercise7-solution}
# Density plot
plot(density(x),
     main = "Density of Binomial Samples (n = 30, p = 0.65)",
     xlab = "Values",
     ylab = "Density",
     lwd = 2)

# Add a rug plot (shows actual observations)
rug(x)
```

```{r exercise7-check}
grade_code()
```

---

### Exercise 8

Use the values generated from `Exercise 6` to Plot a histogram + Density Overlay.

```{r exercise8, exercise=TRUE}
# your code here
```

```{r exercise8-solution}
hist(x,
     probability = TRUE,
     col = "lightgray",
     main = "Histogram with Density Curve",
     xlab = "Values")

lines(density(x), lwd = 2, col = "blue")
```

```{r exercise8-check}
grade_code()
```

---