---
title: "Fundamentals of R"
subtitle: "Statistical Analysis"
author: "Prepared By Clinton Moshe Mwachia"
description: "Built in and external datasets in R"
date: "`r Sys.Date()`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

R provides a powerful set of functions and packages for statistical analysis. In this lesson, we will cover various statistical functions and analysis techniques in R; **minimum, maximum, standard deviation, quartiles, mean, mode, median, summary statistics, range, frequency distribution, and interquartile range**.

## Descriptive Statistics

**Minimum Value**

```{r min}
# create the vector we will work with
vec = c(1,2,3,4,-20,100,230)

# lowest value
min(vec)
```

**Maximum Value**

```{r max}
# highest value
max(vec)
```

**Standard Deviation**

```{r sd}
# Standard Deviation
sd(vec)
```

**Quantiles**

```{r qt}
# quantiles
quantile(vec)
```

**Average**

```{r average}
# average
mean(vec, na.rm = TRUE)
```

**Median**

```{r median}
# median
median(vec, na.rm = TRUE)
```

**Range**

```{r range}
# range
range(vec, na.rm = TRUE)
```

**Interquartile Range**

```{r iqr}
# Interquartile range
IQR(vec, na.rm = TRUE)
```

## Frequency Distribution

A frequency distribution displays the frequency (count) of each unique value in a data set.

```{r fq}
# freaquency
table(vec)
```

## Summary Statistics

Summary statistics provide a comprehensive summary of the central tendency, dispersion, and shape of a data set.

```{r summary}
# summary of vec
summary(vec)
```

---

### Exercise 1

Create a data frame named `df` with the variables `Student`, `Age`, `gender`, `maths`, `pass` and fill it with 20 records.

The variables contain the following data types:

      Student => string
      Age => number
      gender => factor
      maths => number
      pass => boolean
      
Print-out the summary of `df.`

```{r exercise1, exercise=TRUE}
# your code here
```

```{r exercise1-solution}
df <- data.frame(
  Student = c(
    "Alice","Brian","Cathy","David","Eva",
    "Frank","Grace","Henry","Irene","Jack",
    "Kevin","Lily","Mike","Nina","Oscar",
    "Paula","Quinn","Ryan","Sophia","Tom"
  ),
  
  Age = c(
    20,22,19,21,23,
    20,22,21,19,23,
    24,20,22,21,23,
    20,22,21,19,23
  ),
  
  gender = factor(c(
    "Female","Male","Female","Male","Female",
    "Male","Female","Male","Female","Male",
    "Male","Female","Male","Female","Male",
    "Female","Male","Male","Female","Male"
  )),
  
  maths = c(
    85,67,90,55,72,
    88,76,60,95,70,
    66,89,73,81,58,
    92,64,77,86,69
  ),
  
  pass = c(
    TRUE, TRUE, TRUE, FALSE, TRUE,
    TRUE, TRUE, FALSE, TRUE, TRUE,
    TRUE, TRUE, TRUE, TRUE, FALSE,
    TRUE, TRUE, TRUE, TRUE, TRUE
  ),
  
  stringsAsFactors = FALSE
)

summary(df)
```

```{r exercise1-check}
grade_code()
```

---

### Exercise 2

Who is the youngest and oldest student(s)?

```{r exercise2, exercise=TRUE}
# your code here
```

```{r exercise2-solution}
# Youngest student(s)
youngest <- df[df$Age == min(df$Age), ]

# Oldest student(s)
oldest <- df[df$Age == max(df$Age), ]

youngest
oldest
```

```{r exercise2-check}
grade_code()
```

---

### Exercise 3

What percentage of the students are male and female?

```{r exercise3, exercise=TRUE}
# your code here
```

```{r exercise3-solution}
# Count each gender
gender_counts <- table(df$gender)

# Convert to percentages
gender_percent <- prop.table(gender_counts) * 100

gender_percent
```

```{r exercise3-check}
grade_code()
```

---

### Exercise 4

What percentage of students passed the exams?

```{r exercise4, exercise=TRUE}
# your code here
```

```{r exercise4-solution}
# Percentage of students who passed
pass_percent <- mean(df$pass) * 100

pass_percent
```

```{r exercise4-check}
grade_code()
```

---

### Exercise 5

What is the performance range for maths?

```{r exercise5, exercise=TRUE}
# your code here
```

```{r exercise5-solution}
range(df$maths)
```

```{r exercise5-check}
grade_code()
```

---