---
title: "Fundamentals of R"
subtitle: "Conditional Statements"
author: "Prepared By Clinton Moshe Mwachia"
description: "Control Statements in R"
date: "`r Sys.Date()`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Conditional statements are used to  handle decisions in R. They are used to perform certain actions if a certain criteria is met. In this session, we will cover the following :- 

    1. If statements.
    2. If else statements.
    3. Else if statements.
    4. AND and OR.
    5. For loop.
    6. Break Statements.
    7. Next statements.
    
    
## If statements

An if statement is only executed if the condition being tested evaluates to TRUE. The code chunks below will illustrate some of the use cases.

```{r}
num = 20

if(num == 20){
  cat(num)
}
```

## Greater than

```{r}
# check if a number if greater than a given value and print it

# lets define the variables
num = 10

# the if statement
if(num > 10){
  cat(num)
}

# since the num is not greater than 10 (evaluates to FALSE), the code is not
# executed
```

## Equal to

```{r}
# check if num is equal to
if(num == 10){
  cat(num)
}
```

## Not Equal to

```{r}
# check if num is not equal to
if(num != 10){
  cat(num)
}
```

## Greater than or Equal to

```{r}
# check if num is greater than or equal to
if(num >= 10){
  cat(num)
}
```

## Less than or Equal to

```{r}
# check if num is greater than or equal to
if(num <= 10){
  cat(num)
}
```

## If Else Statements

An if else statements is used to handle situations that are not handled by the if statement. For instance, in the if statement, if the num is not equal to 10, the code will not be executed, to solve this we will use the else statement to give the user feedback.

Let use create a new variable, *name*, and use it to illustrate if else statements.

```{r}
# defining name
name = "james bond"

# let check if the name jane doe is equal to name
if(name == "jane doe"){
  cat(name)
} else {
  cat("name is not jane doe")
}
```

Lets change the name from *jane doe* to *james bond*.

```{r}
if(name == "james bond"){
  cat(name)
} else {
  cat("name is not jane doe")
}
```

As seen above, if the if statement does not catch the condition, it is forwarded to the else condition. Lets try the same but with numbers.

```{r}
# declare num2
num2 = 2

# test if the number is greater than 2
if(num2 > 2){
  cat("number is greater than 2")
} else {
  cat("number is less than 2")
}
```

## Else If Statements

Else if statements helps one to handle multiple conditions before executing the else condition.

The code chunk below illustrates how to make use of the else if statements, we will use the name variable.

```{r}
# check if the name is available and give access
name = "james bond"

if(name == "james bond"){
  cat("name exists in the data base")
} else if(name == "Clinton Bond"){
  cat("name clinton exists")
} else {
  cat("name does not exists")
}
```

In the code chunk above, condition 1 is met because the name is james bond.

In the code chunk below, we change the name to clinton Bond, observe the results.

```{r}
# check if the name is available and give access
name = "Clinton Bond"

if(name == "james bond"){
  cat("name exists in the data base")
} else if(name == "Clinton Bond"){
  cat("name clinton exists")
} else {
  cat("name does not exists")
}
```

## AND operator

```{r}
# define the variables
a = 5
b = 7

# check if both are greater than
if(a > 4 & b > 5){
  cat("both a and b are greater than")
} else {
  cat("both a and b are less than")
}

```

## OR operator

```{r}
# check if any is greater than
if(a > 5 | b > 5){
  cat("atleast one greater than")
} else {
  cat("atleast one is less than")
}
```

## For Loop

A for loop is used to iterate over a given sequence and the perform given operations to individual items. The code chunks below illustrate some of the use cases.

## Loop through a vector

```{r}
# define a vector
x = c(1,2,3,4,5)

# print each item of vector x
for (i in x) {
  cat(paste0(i, " ")) # to add spaces between the items
}
```

What if you wanted to multiply the items by a given number? Proceed to the next chunk and find out how to do it or you can also think about first.

```{r}
# multiply by 2 and print each item of vector x 
for (i in x) {
  cat(paste0(i * 2, " ")) # to add spaces between the items
}
```

## Loop through a list

```{r}
# define a list
animals = list("Elephant", "Lion", "Leopard", "Rhino", "Cheetah")

# print each animal
for (animal in animals) {
  cat(paste0(animal, " ")) # to add spaces between the items
}
```

## Break Statement

The break statement is used to break the loop when a certain condition is met. The code below will break from the loop if the animal is Leopard.

```{r}
# define a list
animals = list("Elephant", "Lion", "Leopard", "Rhino", "Cheetah")

# print each animal
for (animal in animals) {
  cat(paste0(animal, " "))
  
  if(animal == "Leopard"){
    break
  }
}
```

## Next statement

The next statement is used to skip an item if a certain condition in met. When the condition being tested evaluates to true, the iteration will move to the next item. The code chunks below will illustrate it's use cases.

```{r}
# define a list
animals = list("Elephant", "Lion", "Leopard", "Rhino", "Cheetah")

# print each animal
for (animal in animals) {
   if(animal == "Leopard"){
    next
   }
   cat(paste0(animal, " "))
}
```

---

## Exercises

### Exercise 1

Create two variables,`x = 10 and y = 20`.

Create a new variable, z, and assign it to the result of y > x. Print the value of z using the `cat()` function

```{r exercise1, exercise=TRUE}
# your code here
```

```{r exercise1-solution}
x = 10
y = 20
z = y > x

cat(z)
```

```{r exercise1-check}
grade_code()
```

---

### Exercise 2

Create a data frame, **Perf** with the columns below;
  *Student = c("James", "Peter", "Andrew", "Mercy",  "Magdalene")*
  *Math = c(89,79,50,40,98)*
  
Create a new column, **Grade**, containing the grades of each student based on the following grading system.

    0 - 30 - D
    31 - 50 - C
    51 - 70 - B
    71 - 80 - B+
    81 - 100 - A

```{r exercise2, exercise=TRUE}
# your code here
```

```{r exercise2-solution}
perf = data.frame(
  Student = c("James", "Peter", "Andrew", "Mercy",  "Magdalene"),
  Math = c(89,79,50,40,98)
)

perf$Grade = ifelse(perf$Math > 80, "A",
                    ifelse(perf$Math > 70, "B+",
                           ifelse(perf$Math > 50, "B",
                                  ifelse(perf$Math > 30, "C", "D"))))

perf
```

```{r exercise2-check}
grade_code()
```

---

### Exercise 3

Using the data frame above, Add a new column, **Pass**, containing a Boolean value on whether a student has passed or not. *The pass mark is 50*.

```{r exercise3, exercise=TRUE}
# your code here
```

```{r exercise3-solution}
perf = data.frame(
  Student = c("James", "Peter", "Andrew", "Mercy",  "Magdalene"),
  Math = c(89,79,50,40,98)
)

perf$Grade = ifelse(perf$Math > 80, "A",
                    ifelse(perf$Math > 70, "B+",
                           ifelse(perf$Math > 50, "B",
                                  ifelse(perf$Math > 30, "C", "D"))))

perf$Pass = ifelse(perf$Math > 50, TRUE, FALSE)

perf
```

```{r exercise3-check}
grade_code()
```

---

### Exercise 4

Create an empty vector, **vec**, of length 10.

Print the contents of *vec*.

```{r exercise4, exercise=TRUE}
# your code here
```

```{r exercise4-solution}
vec = vector(mode="numeric", length = 10)

vec
```

```{r exercise4-check}
grade_code()
```

---

### Exercise 5

Create another vector, **vec2**, containing the data `1,2,3,4,5,6,7,8,9,10`.
Using the appropriate conditional statements, fill up **vec** with twice the value of **vec2**.

```{r exercise5, exercise=TRUE}
# your code here
```

```{r exercise5-solution}
vec = vector(mode="numeric", length = 10)

vec2 = c(1,2,3,4,5,6,7,8,9,10)

for (i in 1:length(vec2)) {
  vec[i] = 2*i
}

vec
```

```{r exercise5-check}
grade_code()
```

---