---
title: "Fundamentals of R"
subtitle: "Hypothesis Testing"
author: "Prepared By Clinton Moshe Mwachia"
description: "Testing Hypothesis in R"
date: "`r Sys.Date()`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Hypothesis testing is a statistical technique used to make inferences about a population based on sample data.

In R, there are several functions and tests available for hypothesis testing.

In this lesson, we will cover the following hypothesis tests: **one-sample t-test, two-sample t-test, normality test, confidence interval, sign test, Wilcoxon signed-rank test, and Kruskal-Wallis rank sum test**.

## One-Sample t-test

The one-sample t-test is used to determine if the mean of a sample significantly differs from a specified value.

The `p-value` obtained from the one-sample t-test determines whether the mean of the sample significantly differs from the specified value (mu = 27).

If the `p-value` is less than the significance level (commonly 0.05), we reject the null hypothesis (H0) and conclude that the mean is significantly different.

```         
  H0: the mean of the data is 27
  H1: the mean is not equal to 27
```

```{r one-sample-ttest}
# Sample Data
data <- c(23, 25, 26, 24, 28, 27, 22, 26, 25, 24)

# One-sample t-test
t.test(data, mu = 27)
```

**Conclusion**

Since the `p-value` is less than 0.05, we reject the null hypothesis, Therefore, we can conclude that the mean of the data is not equal to 27.

## Two-Sample t-test

The two-sample t-test is used to compare the means of two independent samples.

The `p-value` obtained from the two-sample t-test determines whether the means of the two groups significantly differ.

If the `p-value` is less than the significance level (0.05), we reject the null hypothesis and conclude that there is a significant difference between the means of the two groups.

```         
      H0: the mean difference is zero (mean1 = mean2)
      H1: the mean difference is not zero (mean1 != mean2)
```

```{r two-sample-ttest}
# Sample Data
data1 <- c(23, 25, 26, 24, 28)
data2 <- c(27, 22, 26, 25, 24)

# Two-sample t-test
t.test(data1, data2)
```

**Conclusion**

Since the `p-value` is greater than 0.05, we fail to reject the null hypothesis since we do not have enough evidence.

## Normality Test

The normality test is used to assess whether a sample follows a normal distribution.

The Shapiro-Wilk test is commonly used to check for normality. The `p-value` obtained from the test indicates whether the sample deviates significantly from a normal distribution.

If the `p-value` is less than the significance level, we reject the null hypothesis and conclude that the data is not normally distributed.

```         
      H0: the data is normally distributed
      H1: the data is not normally distributed
      
```

```{r shapiro-test}
# Sample Data
data <- rnorm(100)

# Normality test
shapiro.test(data)
```

**Conclusion**

Since the `p-value` is greater than 0.05, we fail to reject the null hypothesis since we do not have enough evidence.

## Wilcoxon Signed-Rank Test

The Wilcoxon signed-rank test is a non-parametric alternative to the paired t-test when the data is not normally distributed.

The `p-value` obtained from the Wilcoxon signed-rank test determines whether there is a significant difference between the paired observations.

If the `p-value` is less than the significance level, we reject the null hypothesis and conclude that there is a significant difference.

```         
  H0: The median of the data of differences between the paired data is zero.
  H1: The median of the data of differences between the paired data is not zero
  
  
```

```{r wilcox-test, warning=FALSE}
# Sample Data
x <- c(23, 25, 26, 24, 28)
y <- c(27, 22, 26, 25, 24)

# Wilcoxon signed-rank test
wilcox.test(x, y, paired = TRUE)
```

**Conclusion**

Since the `p-value` is 1, we fail to reject the null hypothesis and conclude that there is no significant difference between the paired observations.

The test did not find any evidence of a significant difference between the paired observations.

## Kruskal-Wallis Rank Sum Test

The Kruskal-Wallis rank sum test is a non-parametric test used to compare the distributions of two or more independent groups.

The `p-value` obtained from the Kruskal-Wallis test determines whether there is a significant difference between the distributions of the groups.

If the `p-value` is less than the significance level, we reject the null hypothesis and conclude that there is a significant difference.

```         
  H0: The mean ranks of the groups are the same
  H1: The mean ranks of the groups are different
  
```

```{r kruskal-wallis}
# Sample Data
x <- c(23, 25, 26, 24, 28)
y <- c(27, 22, 26, 25, 24)
z <- c(29, 31, 30, 28, 27)

# Kruskal-Wallis test
kruskal.test(list(x, y, z))
```

**Conclusion**

Since the `p-value` is less than 0.05, we reject the null hypothesis and conclude that the mean ranks of the groups are different.

------------------------------------------------------------------------

## Exercises

### Exercise 1

A university collected height data (in cm) from a random sample of male and female students. Is there a statistically significant difference in mean height between male and female students?

Data:
         
    male_heights <- c(172, 180, 175, 178, 182, 177, 169, 181, 176, 174)                female_heights <- c(160, 165, 162, 158, 167,163, 161, 159, 164, 166)

```{r exercise1, exercise=TRUE}
# your code here
```

```{r exercise1-solution}
# Hypotheses
# H₀: μ_male = μ_female
# H₁: μ_male ≠ μ_female
male_heights <- c(172, 180, 175, 178, 182, 177, 169, 181, 176, 174)
female_heights <- c(160, 165, 162, 158, 167, 163, 161, 159, 164, 166)

t.test(male_heights, female_heights)

# Interpretation

# If p < 0.05 → reject H₀ → gender affects height.

# Conclusion

# Male students are likely taller if H₀ is rejected.
```

```{r exercise1-check}
grade_code()
```

------------------------------------------------------------------------

### Exercise 2

The standard recovery time for a disease is 12 days. A new drug is tested on patients. Does the drug significantly reduce recovery time?

Data:

```         
recovery_times <- c(10, 9, 11, 8, 7, 10, 9, 8, 11, 9)
```

```{r exercise2, exercise=TRUE}
# your code here
```

```{r exercise2-solution}
# Hypotheses
# H₀: μ = 12
# H₁: μ < 12
recovery_times <- c(10, 9, 11, 8, 7, 10, 9, 8, 11, 9)

t.test(recovery_times, mu = 12, alternative = "less")

# Interpretation

# p < 0.05 → drug reduces recovery time.

# Conclusion

# Drug is effective if H₀ rejected.
```

```{r exercise2-check}
grade_code()
```

------------------------------------------------------------------------

### Exercise 3

A school tests whether tutoring improves exam scores.

Data:

```         
tutored <- c(78, 85, 82, 88, 90, 84, 79, 87, 91, 86)
not_tutored <- c(65, 70, 68, 72, 66, 69, 71, 67, 64, 73)
```

```{r exercise3, exercise=TRUE}
# your code here
```

```{r exercise3-solution}
# Hypotheses
# H₀: μ₁ = μ₂
# H₁: μ₁ > μ₂
tutored <- c(78, 85, 82, 88, 90, 84, 79, 87, 91, 86)
not_tutored <- c(65, 70, 68, 72, 66, 69, 71, 67, 64, 73)

t.test(tutored, not_tutored, alternative = "greater")

# Interpretation

# p < 0.05 → tutoring improves scores.

# Conclusion

# Tutoring has a positive effect.
```

```{r exercise3-check}
grade_code()
```

------------------------------------------------------------------------

### Exercise 4

A company’s average weekly sales are KES 50,000. After a new marketing strategy, sales are recorded. Did sales increase?

Data:

```         
sales <- c(52000, 55000, 53000, 60000, 58000, 57000, 56000, 59000)
```

```{r exercise4, exercise=TRUE}
# your code here
```

```{r exercise4-solution}
# Hypotheses
# H₀: μ = 50000
# H₁: μ > 50000
sales <- c(52000, 55000, 53000, 60000, 58000, 57000, 56000, 59000)

t.test(sales, mu = 50000, alternative = "greater")

# Interpretation

# p < 0.05 → sales increased significantly.

# Conclusion

# Strategy is effective.
```

```{r exercise4-check}
grade_code()
```

------------------------------------------------------------------------

### Exercise 5

Two website designs are tested for response time (seconds). Which one is faster?

Data:

```         
design_old <- c(2.5, 2.7, 2.6, 2.8, 2.4, 2.6, 2.7, 2.5)
design_new <- c(2.1, 2.0, 1.9, 2.2, 2.1, 2.0, 1.8, 2.1)
```

```{r exercise5, exercise=TRUE}
# your code here
```

```{r exercise5-solution}
# Hypotheses
# H₀: μ_old = μ_new
# H₁: μ_new < μ_old
design_old <- c(2.5, 2.7, 2.6, 2.8, 2.4, 2.6, 2.7, 2.5)
design_new <- c(2.1, 2.0, 1.9, 2.2, 2.1, 2.0, 1.8, 2.1)

t.test(design_new, design_old, alternative = "less")

# Interpretation

# p < 0.05 → new design is faster.

# Conclusion

# New design improves performance.
```

```{r exercise5-check}
grade_code()
```

------------------------------------------------------------------------
