---
title: "Fundamentals of R"
subtitle: "Functions"
author: "Prepared By Clinton Moshe Mwachia"
description: "Control Statements in R"
date: "`r Sys.Date()`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

A function in R is a block of code that performs certain tasks and can only run when it is called. Functions are normally created and stored into libraries. In this session, we will learn about the following: -

      1. Built on functions.
      2. User defined functions.
      
## Built in Functions

These are functions that can be found in the default R work-space. They come with the R and RStudio ecosystem and can be accessed without installing any libraries. Some of the built functions in R are `mean`, `mode`,`median`, `sum` and many more, The code chunks below will show some of the built in functions.

**Absolute Value**

```{r}
# get the absolute value of a number
abs(-10)
```

**Sum Numbers**

```{r}
# sum of 10 numbers
# initialize the sum
total = 0 
# loop through 10 numbers and sum them
for (i in 1:10) {
  total = total + i
}

# print the total
cat(total)
```

**Logarithm Of a Number**

```{r}
# Logarithm of a number
log(2)
```

**Split a String by space**

```{r}
# Split a string by space

# define the string
name = "James clinton"

# split the string by space
strsplit(name, " ")
```

**Calculate Average of Numbers**

```{r}
# the mean of 2 numbers
mean(c(1,3))
```

## User Defined Functions

These a functions created by the user to perform specific tasks. They are mainly created so they can be re-used.

**Function with arguments**

```{r}
# defining the function
Hello <- function(name){
  cat(paste0("Hello, ", name))
}

# test case 1
Hello("clinton")

# test case 2
Hello("James Peter")
```

**Function without arguments**

```{r}
# defining the function
Hello2 <- function(){
  cat("Hello, Clinton")
}

# test case 1
Hello2()
```

**Function with default arguments**

```{r}
## Function with default arguments
# defining the function
Hello3 <- function(name = "Clinton Moshe"){
  cat(paste0("\nHello, ", name))
}

# test case 1
Hello3()

# test case 2
Hello3("John Doe")
```

*NOTE: IF YOU WANT THE FUNCTION TO RETURN A VALUE,USE THE RETURN STATEMENT AS ILLUSTRATED BELOW*

**The function will return the sum of 20 numbers.**

```{r}
# defining the number
sum_20 <- function(){
  # initialize the total
  tot = 0
  # loop through 20 numbers and sum
  for (i in 1:20) {
    tot = tot + i
  }
  # return the value
  return(tot)
}

# calling the function
sum_20()
```

