---
title: "Beginner: Fundamentals of R"
subtitle: "Data Structures"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: learnr::tutorial
runtime: shiny_prerendered
tutorial: true
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

# 👋 Introduction

Data structures define how data is stored and organized in R.

In this lesson, you will learn:

* Vectors
* Lists
* Matrices
* Data Frames
* Bonus: Variables, Comments

---

# 🔢 VECTORS

A vector stores elements of the same type.

```{r}
vec <- c(1,2,3,4)
vec
```

---

## 🧠 Exercise 1: Create a Vector

Create a vector called `animals` with 5 animals.

```{r exercise1, exercise=TRUE}
# your code here
```

```{r exercise1-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")
```

```{r exercise1-check}
grade_code("CORRECT","INCORRECT")
```

---

## 🧠 Exercise 2: Vector Length

Find the length of `animals`.

```{r exercise2, exercise=TRUE}
# your code here
```

```{r exercise2-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")

length(animals)
```

```{r exercise2-check}
grade_code()
```

---

## 🧠 Exercise 3: Indexing

Extract the 1st, 3rd and 5th animals.

```{r exercise3, exercise=TRUE}
# your code here
```

```{r exercise3-solution}
animals[c(1,3,5)]
```

```{r grade3}
grade_this({
  animals[c(1,3,5)]
},
pass_if_equal(c("Lion","Buffalo","Rhino"),
              message = "✅ Correct!"))
```

---

# 📋 LISTS

Lists can store different data types.

```{r}
my_list <- list(10, TRUE, "R")
my_list
```

---

## 🧠 Exercise 4: Create a List

Create a list `my_list` using `animals`.

```{r exercise4, exercise=TRUE}
# your code here
```

```{r exercise4-solution}
my_list <- list(animals)
```

```{r grade4}
if (!exists("my_list")) {
  fail("❌ Create my_list")
} else {
  pass("✅ List created!")
}
```

---

# 🔢 MATRICES

Matrices are 2D structures.

```{r}
A <- matrix(c(1,2,3,4,5,6), nrow = 2)
A
```

---

## 🧠 Exercise 5: Matrix Dimensions

Find dimensions of matrix `A`.

```{r exercise5, exercise=TRUE}
# your code here
```

```{r exercise5-solution}
dim(A)
```

```{r grade5}
grade_code({
  dim(A)
})
```

---

## 🧠 Exercise 6: Matrix Addition

Compute A + A

```{r exercise6, exercise=TRUE}
# your code here
```

```{r exercise6-solution}
A + A
```

```{r grade6}
grade_this({
  A + A
},
pass_if_equal(A + A,
              message = "✅ Correct! You successfully added matrix A to itself."))
```

---

# 📊 DATA FRAMES

Data frames store tabular data.

```{r}
df <- data.frame(one=c(10.5,11,23,4))
df
```

---

## 🧠 Exercise 7: Summary Stats

Get summary of df

```{r exercise7, exercise=TRUE}
# your code here
```

```{r exercise7-solution}
summary(df)
```

```{r grade7}
grade_code({
  summary(df)
})
```

---

## 🧠 Exercise 8: Add Column

Add a new column called `two`

```{r exercise8, exercise=TRUE}
# your code here
```

```{r exercise8-solution}
df$two <- c(1,3,5,7)
```

```{r grade8}
if (!"two" %in% colnames(df)) {
  fail("❌ Add column 'two'")
} else {
  pass("✅ Column added!")
}
```

---

# 🧠 BONUS

## Comments

```{r}
# This is a comment
```

---

## Variables

```{r}
x <- 10
y = 20
```

---

# 🏁 FINAL CHALLENGE (Assignment)

## 🧠 Exercise 9: Full Task

Create:

* animals vector
* marks vector
* list combining both

Then return length of list

```{r exercise9, exercise=TRUE}
# your code here
```

```{r exercise9-solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")
marks <- c(70,80,90,85,60)

my_list <- list(animals, marks)

length(my_list)
```

```{r grade9}
if (!exists("my_list")) {
  fail("❌ Create my_list")
} else if (length(my_list) != 2) {
  fail("❌ It should contain two items")
} else {
  pass("✅ Excellent work! 🎉")
}
```

---

# 🎯 Summary

You learned:

* Vectors
* Lists
* Matrices
* Data Frames
* Basic operations

---

# 🚀 Next Lesson

👉 Continue to: **Data Structures Assignment / Advanced Practice**
