---
title: "Beginner: Fundamentals of R"
subtitle: "2.0 Data Structures"
author: "Prepared By Clinton Moshe Mwachia"
date: "`r Sys.Date()`"
output: learnr::tutorial
runtime: shiny_prerendered
description: "Learn R data structures interactively with exercises and instant feedback"
----------------------------------------------------------------------------------------

```{r setup, include=FALSE}
library(learnr)
library(gradethis)

gradethis::gradethis_setup()
knitr::opts_chunk$set(echo = TRUE)
```

# 👋 Introduction

Data structures define how data is stored and organized in R.

In this lesson, you will learn:

* Vectors
* Lists
* Matrices
* Data Frames
* Bonus: Variables, Comments

---

# 🔢 VECTORS

A vector stores elements of the same type.

```{r}
vec <- c(1,2,3,4)
vec
```

---

## 🧠 Exercise 1: Create a Vector

Create a vector called `animals` with 5 animals.

```{r exercise=TRUE}
# your code here
```

```{r solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")
```

```{r grade}
if (!exists("animals")) {
  fail("❌ Create a variable called animals")
} else if (length(animals) != 5) {
  fail("❌ It should contain 5 animals")
} else {
  pass("✅ Good job!")
}
```

---

## 🧠 Exercise 2: Vector Length

Find the length of `animals`.

```{r exercise=TRUE}
# your code here
```

```{r solution}
length(animals)
```

```{r grade}
grade_code({
  length(animals)
})
```

---

## 🧠 Exercise 3: Indexing

Extract the 1st, 3rd and 5th animals.

```{r exercise=TRUE}
# your code here
```

```{r solution}
animals[c(1,3,5)]
```

```{r grade}
grade_result(
  pass_if = length(.result) == 3,
  correct = "✅ Correct!",
  fail = "❌ Use indexing like c(1,3,5)"
)
```

---

# 📋 LISTS

Lists can store different data types.

```{r}
my_list <- list(10, TRUE, "R")
my_list
```

---

## 🧠 Exercise 4: Create a List

Create a list `my_list` using `animals`.

```{r exercise=TRUE}
# your code here
```

```{r solution}
my_list <- list(animals)
```

```{r grade}
if (!exists("my_list")) {
  fail("❌ Create my_list")
} else {
  pass("✅ List created!")
}
```

---

# 🔢 MATRICES

Matrices are 2D structures.

```{r}
A <- matrix(c(1,2,3,4,5,6), nrow = 2)
A
```

---

## 🧠 Exercise 5: Matrix Dimensions

Find dimensions of matrix `A`.

```{r exercise=TRUE}
# your code here
```

```{r solution}
dim(A)
```

```{r grade}
grade_code({
  dim(A)
})
```

---

## 🧠 Exercise 6: Matrix Addition

Compute A + A

```{r exercise=TRUE}
# your code here
```

```{r solution}
A + A
```

```{r grade}
grade_result(
  pass_if = all(.result == (A + A)),
  correct = "✅ Correct!",
  fail = "❌ Add matrix A to itself"
)
```

---

# 📊 DATA FRAMES

Data frames store tabular data.

```{r}
df <- data.frame(one=c(10.5,11,23,4))
df
```

---

## 🧠 Exercise 7: Summary Stats

Get summary of df

```{r exercise=TRUE}
# your code here
```

```{r solution}
summary(df)
```

```{r grade}
grade_code({
  summary(df)
})
```

---

## 🧠 Exercise 8: Add Column

Add a new column called `two`

```{r exercise=TRUE}
# your code here
```

```{r solution}
df$two <- c(1,3,5,7)
```

```{r grade}
if (!"two" %in% colnames(df)) {
  fail("❌ Add column 'two'")
} else {
  pass("✅ Column added!")
}
```

---

# 🧠 BONUS

## Comments

```{r}
# This is a comment
```

---

## Variables

```{r}
x <- 10
y = 20
```

---

# 🏁 FINAL CHALLENGE (Assignment)

## 🧠 Exercise 9: Full Task

Create:

* animals vector
* marks vector
* list combining both

Then return length of list

```{r exercise=TRUE}
# your code here
```

```{r solution}
animals <- c("Lion","Elephant","Buffalo","Leopard","Rhino")
marks <- c(70,80,90,85,60)

my_list <- list(animals, marks)

length(my_list)
```

```{r grade}
if (!exists("my_list")) {
  fail("❌ Create my_list")
} else if (length(my_list) != 2) {
  fail("❌ It should contain two items")
} else {
  pass("✅ Excellent work! 🎉")
}
```

---

# 🎯 Summary

You learned:

* Vectors
* Lists
* Matrices
* Data Frames
* Basic operations

---

# 🚀 Next Lesson

👉 Continue to: **Data Structures Assignment / Advanced Practice**
